Efficient pedigree recording for fast population genetics simulation

### 2018-01-12 forwards_paper.tex
Light editing pass.

\begin{abstract}
    In this paper we describe how to
    efficiently record the entire genetic history of a population
    during a forwards-time, individual-based population genetics simulation.
    This dramatically reduces the computational burden of tracking individual genomes
    by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording history as a \emph{succinct tree sequence} as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    We implement the method in two popular forwards-time simulation frameworks,
    which speeds up large, whole-genome simulations by one to two orders of magnitude.
    In addition to speed, the method has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
    can be stored in $O(N \log N + M)$ space, making it feasible to store a simulation's entire final generation (and its' history)
    rather than only a subset.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation of deep history.
    To make these possible,
    we introduce the \msprime{} Tables API, which allows for the efficient interchange
    of tree sequences between modular program components,
    and show how to quickly simplify a tree sequence by removing history irrelevant to a given set of genomes.
\end{abstract}

@@

In this paper, we describe a storage method for \emph{succinct tree sequences}
and, hence, genome sequence, as well as an algorithm for simplifying these.
The data structure is \emph{succinct} in the sense that its' space usage is close to optimal,
while still allowing efficient retrieval of information (see, e.g.,
\citet{gog2014theory}).
Both are implemented as improvements to the algorithmic tools and data structures in \msprime{}.
We also describe how these tools can efficiently record,
and later process, the embellished population pedigree from a forwards-time simulation.
While providing substantial savings in computational time and space, our methods provide much
more information than simply simulating the genomes -- the tree sequence
encodes all marginal genealogies of individuals living at the end of the simulation.
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.

@@

The strategy described above is only of interest if it is computationally feasible.
Therefore,
we begin by benchmarking the performance improvement achieved by this method,
implemented using the forwards-time simulation library \fwdpp{} \citep{fwdpp}
and tree sequence tools implemented in \msprime{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences succinctly in several \emph{tables};
(b) an algorithm to record these tables on the fly during a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
Finally, we analyze the run time and space complexity of our general-purpose method.

@@

% When tracking neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for these simulations).
Pedigree tracking dramatically reduced runtimes, as shown in Figure~\ref{fig:runtimes_selection},
producing a relative speedup of up to $\approx 50$ fold relative to standard simulations that track neutral mutations
(Figure~\ref{fig:relative_speedup_selection}).
Pedigree tracking results in greater relative speedups for larger $N$
and we observe increasing relative speedups as $4Nr$ increases for a given $N$
(Figure~\ref{fig:relative_speedup_selection}).
Importantly, runtimes are approximately linear in region size $\rho$ when pedigree tracking
(partially obscured by the log scale of the horizontal axis in Figure~\ref{fig:runtimes_selection}).
%% don't know what an 'edge' is yet
% The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
% each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
In a more limited set of neutral simulations we found the same qualitative behavior,
and a numerically larger speedup to pedigree tracking (see Appendix~\ref{ss:timing_nosel}).

@@

@@

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
records two things: its position on the genome and an ancestral state.
For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2.5 with ancestral state `A' and the other at position 7.5 with ancestral state `G'.
If no mutations occur at a given site, all nodes inherit the ancestral state.

Each mutation records three things: the site at which it occurs,
the first node to inherit the mutation, and the derived state.
Thus, all nodes below the mutation's node in the tree will inherit this state,
unless further mutations are encountered.
Three mutations are shown in Figure~\ref{fig:example_tree_sequence},
illustrated by red stars.
The first site, in the left-hand tree,
has a single mutation, which results in node $2$ inheriting the state `T'.
The second site, in the right hand tree, has two mutations:
one occurring over node $3$ changing the state to `C',
and a back mutation over node $1$ changing the state to `G'.

@@

\paragraph{Flexibility.}
To demonstrate the flexibility provided by the Tables API and provide an
implementation that decouples forward simulation internals from transfer of data
to \msprime, we also implemented a version of the simulations described in
``Simulation benchmarks'' separately in Python, described in Appendix~\ref{ss:simupop}.
This proof-of-concept implementation, the simulation engine (we use \simupop{}, \citet{simupop})
invokes callbacks at critical points of the simulation, and we infer nodes and edges
from the information that is provided. Rows are appended to the tables
one-by-one, and the tables are periodically sorted and simplified to control
memory usage.
Benchmarking results from this implementation are shown (alongside results from \fwdpp{})
for simulations without selection in Appendix~\ref{ss:timing_nosel}:
a relatively modest speedup of around $5 \times$ is achieved, likely due to increased overhead.

@@

\algstep{W3.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, N - 1\})$,
    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform([0, L))$.
}

@@

This algorithm records only topological information about the simulated genealogies,
but it is straightforward to add mutational information.
Mutations that occur during the simulation can be recorded
by simply recording the node in which they first occur, the derived state,
and (if not already present) the genomic position of the site at which it occurs.
This allows selected mutations, that the forwards time simulation must generate,
to be recorded in the tree sequence.
Neutral mutations can be generated after the simulation has completed, thus
avoiding the cost of generating the many mutations that are lost in the population.
This is straightforward to do because we have access to the marginal genealogies.

@@

It is desirable for many reasons to remove redundant information from a tree sequence,
such as those produced by a forwards-time simulation.
To formalize this:
suppose that we are only interested in a subset of the nodes of a tree sequence
(which we refer to as our `samples'),
and wish to reduce this input tree sequence
to the smallest one that still completely describes the history of the specified samples,
having the following properties:
\begin{enumerate}

@@

Any simulation scheme that records data into tables,
as Algorithm~\algref{W} does,
has its genealogical history available at any time as a tree sequence.
This has two additional advantages:
First, simplification can be run periodically through the simulation,
if we take the set of samples to be the entire currently alive population.
This is important in practice as it keeps memory usage from growing linearly (and quickly) with time.
Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime,
providing an easy, efficient way to specify prior history.
A natural question is now: how often should simplification occur?
Limited testing (described in Appendix~\ref{ss:gcinterval})
found that different simplification intervals affect run times by approximately
25\%, with the lowest run time occurring when simplifying every $10^3$ generations.
Thus, there is a memory-versus-speed tradeoff
-- simplifying more often would keep fewer extinct nodes and edges in memory.
The run-time costs associated
with simplifying have to do with preparing the data to be sent to the simplification algorithm, rather than with the
algorithm itself, which we show in the next section has linear time complexity.

@@

The tree sequences produced by default by this method
are very compact, storing genotype \emph{and} genealogical information
in a small fraction of the space taken by a compressed VCF file.
The format also allows highly efficient processing for downstream analysis.
Efficient processing is possible because many statistics of interest for population genetics
are naturally expressed in terms of tree topologies,
and so can be quickly computed from the trees underlying the tree sequence format.
For example, pairwise nucleotide diversity $\pi$, is the average density of
differences between sequences in the sample.
To compute this directly from sequence data at $m$ sites in $n$ samples
requires computing allele frequencies, taking $O(nm)$ operations.
By using the locations of the mutations on the marginal trees,
and the fact that these are correlated,
sequential tree algorithms similar to those in~\citep{kelleher2016efficient}
can do this in roughly $O(n + m \log n)$ operations.
The \msprime{} API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, which took about 0.7 seconds when
applied to an example simulation of 100 megabases of human-like
sequence for 200,000 samples (about 500K sites). The corresponding
numeric genotype matrix required about 95GiB of RAM, and
calculating $\pi$ took about 66 seconds with NumPy.

@@

\paragraph{Relationship to the ARG}
The idea of a tree sequence is closely related to the \emph{ancestral recombination graph},
or {ARG} \citep{griffiths1991two,griffiths1997ancestral},
which also describes the embellished pedigree.
The ARG has been the subject of substantial study
under the assumptions of coalescent
theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
However, the properties of the ARG as a computational structure have not
been studied and, despite several efforts to standardise a common
format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper
ARGs are rarely used in practise.
In contrast, the algorithmic properties of tree sequence
algorithms have been explored in detail~\citep{kelleher2016efficient},
contributing substantially to the efficiency of the \msprime{} coalescent simulator.

\paragraph{Bayesian phylogenetics}
The methods described here for efficiently storing tree sequences may prove useful in other fields.
For instance, the large numbers of large, correlated trees produced by
MCMC samplers used in Bayesian phylogenetics \citep[e.g.,][]{drummond2012bayesian}
might be compactly stored as a tree sequence,
which would allow highly efficient computation of properties of the posterior distribution.

\paragraph{Next steps.}
While our methods result in significant performance
gains over current simulation methods, they are far from optimal. Because
forwards time simulation is highly CPU intensive, most simulation engines
are written in \justc{} or \cpp{} out of necessity. Therefore, the simplest and most
efficient approach would be to call the \msprime{} \justc{} library directly as part of
the core simulation logic, in the manner outlined in Algorithm \algref{W}. If designed
in this way, the simulation engine would not need to keep copies of the node and edge information.
To facilitate this usage, we plan to release a standalone, embeddable \justc{} library encompassing the core
tree sequence functionality, which should make recording and
outputting tree sequences from any \justc{}/\cpp{} code straightforward.

>>>>

\begin{abstract}
    In this paper we describe how to
    efficiently record the entire genetic history of a population
    during a forwards-time, individual-based population genetics simulation.
    This dramatically reduces the computational burden of tracking individual genomes
    by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording history as a \emph{succinct tree sequence} as introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    We implement the method in two popular forwards-time simulation frameworks,
    which speeds up large, whole-genome simulations by one to two orders of magnitude.
    In addition to speed, the method has several advantages:
    (1) All marginal genealogies of the simulated individuals are recorded, rather than just genotypes.
    (2) A population of $N$ individuals with $M$ polymorphic sites
    can be stored in $O(N \log N + M)$ space, making it feasible to store a simulation's entire final generation (and its' history)
    rather than only a subset.
    (3) A simulation can easily be initialized with a more efficient coalescent simulation of deep history.
    To make these possible,
    we introduce the \msprime{} Tables API, which allows for the efficient interchange
    of tree sequences between modular program components,
    and provide an algorithm to quickly simplify a tree sequence by removing history irrelevant to a given set of genomes.
\end{abstract}

@@

In this paper, we describe a storage method for \emph{succinct tree sequences}
(and hence, genome sequence) as well as an algorithm for simplifying these.
The data structure is \emph{succinct} in the sense that its' space usage is close to optimal,
while still allowing efficient retrieval of information (see, e.g.,
\citet{gog2014theory}).
The storage method and simplification algorithm are both implemented
as improvements to the algorithmic tools and data structures in \msprime{}.
We also describe how these tools can efficiently record,
and later process, the embellished population pedigree from a forwards-time simulation.
While providing substantial savings in computational time and space, our methods provide much
more information than simply simulating the genomes -- the tree sequence
encodes all marginal genealogies of individuals living at the end of the simulation.
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.

@@

The strategy described above is only of interest if it is computationally feasible.
Therefore,
we begin by benchmarking the performance improvement achieved by this method,
implemented using the forwards-time simulation library \fwdpp{} \citep{fwdpp}
and tree sequence tools implemented in \msprime{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences efficiently in several \emph{tables};
(b) an algorithm to record these tables during a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
Finally, we analyze the run time and space complexity of our general-purpose method.

@@

% When tracking neutral mutations (instead of the pedigree), run times increase dramatically with increasing region size ($4Nr = 4Nu$ for these simulations).
Pedigree tracking dramatically reduced runtimes, as shown in Figure~\ref{fig:runtimes_selection},
producing a relative speedup of up to $\approx 50$ fold relative to standard simulations that track neutral mutations
(Figure~\ref{fig:relative_speedup_selection}).
Pedigree tracking results in greater relative speedups for larger $N$
and we observe increasing relative speedups as $4Nr$ increases for a given $N$
(Figure~\ref{fig:relative_speedup_selection}).
Importantly, runtimes are approximately linear in region size $\rho$ when pedigree tracking
(partially obscured by the log scale of the horizontal axis in Figure~\ref{fig:runtimes_selection}).
%% don't know what an 'edge' is yet
% The reason for this behavior is that, when tracking pedigrees, the simulations are generating a constant number of new edges on average
% each generation as the number of new recombination breakpoints is Poisson distributed with expectation $\rho$.
In a more limited set of neutral simulations we found the same qualitative behavior,
and a numerically larger speedup by using pedigree tracking (see Appendix~\ref{ss:timing_nosel}).

@@

The idea of a tree sequence is closely related to the \emph{ancestral recombination graph},
or {ARG} \citep{griffiths1991two,griffiths1997ancestral},
which also describes the embellished pedigree.
The ARG has been the subject of substantial study
under the assumptions of coalescent
theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
However, the properties of the ARG as a computational structure have not
been studied and, despite several efforts to standardise a common
format~\citep{morin2006netgen,mcgill2013graphml}, % TODO check these refs against others in msprime paper
ARGs are rarely used in practise.
In contrast, the algorithmic properties of tree sequence
algorithms have been explored in detail~\citep{kelleher2016efficient},
contributing substantially to the efficiency of the \msprime{} coalescent simulator.

@@

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
records two things: its position on the genome and an ancestral state.
For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2.5 with ancestral state `A' and the other at position 7.5 with ancestral state `G'.
If no mutations occur at a given site, all nodes inherit the ancestral state.
Each mutation records three things: the site at which it occurs,
the first node to inherit the mutation, and the derived state.
Thus, all nodes below the mutation's node in the tree will inherit this state,
unless further mutations are encountered.
Three mutations are shown in Figure~\ref{fig:example_tree_sequence},
illustrated by red stars.
The first site, in the left-hand tree,
has a single mutation, which results in node $2$ inheriting the state `T'.
The second site, in the right hand tree, has two mutations:
one occurring over node $3$ changing the state to `C',
and a back mutation over node $1$ changing the state to `G'.

@@

\paragraph{Flexibility.}
To demonstrate the flexibility provided by the Tables API and provide an
implementation that decouples forward simulation internals from transfer of data
to \msprime, we also implemented a version of the simulations described in
``Simulation benchmarks'' separately in Python, described in Appendix~\ref{ss:simupop}.
In this proof-of-concept implementation, the simulation engine (we use \simupop{}, \citet{simupop})
invokes callbacks at critical points of the simulation, and we infer nodes and edges
from the information that is provided. Rows are appended to the tables
one-by-one, and the tables are periodically sorted and simplified to control
memory usage.
Benchmarking results from this implementation are shown (alongside results from \fwdpp{})
for simulations without selection in Appendix~\ref{ss:timing_nosel}:
a relatively modest speedup of around $5 \times$ is achieved, likely due to increased overhead.

@@

\algstep{W3.}{Choose parents.}{Set $a \leftarrow \randomuniform(\{0, \dots, N - 1\})$,
    $b \leftarrow \randomuniform(\{0, \dots, N - 1\})$ and $x \leftarrow \randomuniform((0, L))$.
}

@@

This algorithm records only topological information about the simulated genealogies,
but it is straightforward to add mutational information.
Mutations that occur during the simulation can be recorded
by simply storing the node in which they first occur, the derived state,
and (if not already present) the genomic position of the site at which it occurs.
This allows selected mutations, that the forwards time simulation must generate,
to be recorded in the tree sequence.
Neutral mutations can be generated after the simulation has completed, thus
avoiding the cost of generating the many mutations that are lost in the population.
This is straightforward to do because we have access to the marginal genealogies.

@@

It is desirable for many reasons to remove redundant information from a tree sequence.
To formalize this:
suppose that we are only interested in a subset of the nodes of a tree sequence
(which we refer to as our `samples'),
and wish to reduce this input tree sequence
to the smallest one that still completely describes the history of the specified samples,
having the following properties:
\begin{enumerate}

@@

Any simulation scheme that records data into tables,
as Algorithm~\algref{W} does,
has its genealogical history available at any time as a tree sequence.
This has two additional advantages:
First, simplification can be run periodically through the simulation,
if we take the set of samples to be the entire currently alive population.
This is important in practice as it keeps memory usage from growing linearly (and quickly) with time.
Second, the simulation can
be \emph{begun} with a tree sequence produced by some other method -- for
instance, by a coalescent simulation with \msprime,
providing an easy, efficient way to specify prior history.
A natural question is now: how often should simplification occur?
Limited testing (described in Appendix~\ref{ss:gcinterval})
found that different simplification intervals affect run times by approximately
25\%, with the lowest run time occurring when simplifying every $10^3$ generations.
Thus, there is a memory-versus-speed tradeoff
-- simplifying more often would keep fewer extinct nodes and edges in memory.

@@

The tree sequences produced by default by this method
are very compact, storing genotype \emph{and} genealogical information
in a small fraction of the space taken by a compressed VCF file.
The format also allows highly efficient processing for downstream analysis.
Efficient processing is possible because many statistics of interest for population genetics
are naturally expressed in terms of tree topologies,
and so can be quickly computed from the trees underlying the tree sequence format.
For example, pairwise nucleotide diversity $\pi$, is the average density of
differences between sequences in the sample.
To compute this directly from sequence data at $m$ sites in $n$ samples
requires computing allele frequencies, taking $O(nm)$ operations.
By using the locations of the mutations on the marginal trees,
and the fact that these are correlated,
sequential tree algorithms similar to those in~\citep{kelleher2016efficient}
can do this in roughly $O(n + m + t)$ operations, where
$t$ is the number of distinct trees.
The \msprime{} API provides a method to compute $\pi$ among arbitrary subsets of the
samples in a tree sequence, which took about 0.7 seconds when
applied to an example simulation of 100 megabases of human-like
sequence for 200,000 samples (about 500K sites). The corresponding
numeric genotype matrix required about 95GiB of RAM, and
calculating $\pi$ took about 66 seconds with NumPy.

@@

\paragraph{Bayesian phylogenetics}
The methods described here for efficiently storing tree sequences may prove useful in other fields.
We have focused on the interpretation of tree sequences as the outcome of the process
of recombination, but in principle, we can efficiently encode any
sequence of trees which differ by subtree-prune-and-regraft
(SPR) operations. Since each SPR operation requires a constant amount of space to encode, the total
space required is $O(n + t)$ for $t$ trees with $n$ leaves~\citep{kelleher2016efficient}.
For instance, the large numbers of large, correlated trees produced by
MCMC samplers used in Bayesian phylogenetics \citep[e.g.,][]{drummond2012bayesian}
might be compactly stored as a tree sequence,
which would then allow highly efficient computation of properties of the posterior distribution.

% JK: This needs more context. Probably not worth putting in here.
% \paragraph{Next steps.}
% While our methods result in significant performance
% gains over current simulation methods, they are far from optimal. Because
% forwards time simulation is highly CPU intensive, most simulation engines
% are written in \justc{} or \cpp{} out of necessity. Therefore, the simplest and most
% efficient approach would be to call the \msprime{} \justc{} library directly as part of
% the core simulation logic, in the manner outlined in Algorithm \algref{W}. If designed
% in this way, the simulation engine would not need to keep copies of the node and edge information.
% To facilitate this usage, we plan to release a standalone, embeddable \justc{} library encompassing the core
% tree sequence functionality, which should make recording and
% outputting tree sequences from any \justc{}/\cpp{} code straightforward.
